---
output: github_document
editor_options: 
  chunk_output_type: console
---

[![Coverage status](https://codecov.io/gh/rtaph/outcomerate/branch/master/graph/badge.svg)](https://codecov.io/github/rtaph/outcomerate?branch=master) [![Travis build status](https://travis-ci.org/rtaph/outcomerate.svg?branch=master)](https://travis-ci.org/rtaph/outcomerate) [![Ropensci status](https://badges.ropensci.org/213_status.svg)](https://github.com/ropensci/onboarding/issues/213)

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# outcomerate

`outcomerate` is a lightweight R package that implements the standard outcome rates for surveys, as defined in the [Standard Definitions](https://www.aapor.org/Standards-Ethics/Standard-Definitions-(1).aspx) of the American Association of Public Opinion Research (AAPOR).

Although the mathematical formulas are straightforward, it can get tedious and repetitive calculating all the rates by hand. The formulas are similar to one another and so it is also dangerously easy to make a clerical mistake. The `outcomerate` package simplifies the analytically workflow by defining all formulas as a collection of functions.


## Installation

The package is under development and available via github:

``` r
install.packages("devtools")
devtools::install_github("rtaph/outcomerate")
```

## Example

Let's say you try to survey 11 people. After finishing the fieldwork, you tabulate all your attempts into a table of disposition outcomes:

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(forcats)
library(knitr)
options(digits = 2)


x <- c("I", "P", "R", "NC", "O", "UH",  "I",  "I", "I", "P", "UO")
data.frame(code = x) %>%
  mutate_at(vars(code), funs(fct_inorder)) %>%
  mutate(disposition = fct_recode(x, `Complete interview` = "I", 
                                  `Partial interview` = "P",
                                  `Refusal and break-off` = "R",
                                  `Non-contact` = "NC",
                                  `Other` = "O",
                                  `Unknown if household` = "UH",
                                  `Unknown, other` = "UO")) %>%
  count(code, disposition) %>%
  kable()
```

Using this table, you may wish to report some of the common survey outcome rates, such as:

* Response Rate
* Cooperation Rate
* Refusal Rate
* Contact Rate
* Location Rate

Most of these rates come under a number of variants, having definitions that are standardized by AAPOR. The `outcomerate` function lets your calculate these rates seamlessly:

```{r example}
# load package
library(outcomerate)

# set counts per disposition code (needs to be a named vector)
freq <- c(I = 4, P = 2, R = 1, NC = 1, O = 1, UH = 1, UO = 1)

# calculate rates, assuming 50% of unknown cases are elligble
outcomerate(freq, e = 0.5)
```

Dispositions do not always come in a tabulated format. Survey analysts often work with microdata directly, where each row represents an interview. The `outcomerate` package allows you to obtain rates using such a format as well:

```{r example2}
# define a vector of dispositions
x <- c("I", "P", "I", "UO", "R", "I", "NC", "I", "O", "P", "UH")

# calculate desired rates
outcomerate(x, rate = c("RR2", "CON1"))

# obtain a weighted rate
w <- c(rep(1.3, 6), rep(2.5, 5))
outcomerate(x, weight = w, rate = c("RR2", "CON1"))
```
